#Script to perform differential expression analysis using DESeq2

#Load libraries
```{r}
library(DESeq2)
library(tximport)
library(tidyverse)
library(dplyr)
library(biomaRt)
library(writexl)
```

# Step 1: Read in sample and gene metadata

```{r}

metadata <- read.csv("./metadata.csv")

mart <- useEnsembl(biomart="ensembl", dataset="hsapiens_gene_ensembl")

t2g <- getBM(
  attributes = c("ensembl_transcript_id", "external_gene_name", "transcript_biotype"),
  filters = "transcript_biotype",
  values = "protein_coding",
  mart = mart
)

```

# Step 2: Read in gene quantification data to a DESeq2 object

```{r}

file_names = metadata$FileName

file_names = file.path("./quants/", file_names)

names(file_names) = metadata$FileName

tx_data = tximport(file_names,
                   type = c("salmon"),
                   tx2gene = t2g,
                   abundanceCol = "TPM",
                   countsCol = "NumReads",
                   lengthCol = "EffectiveLength",
                   ignoreTxVersion = T,
                   ignoreAfterBar = T)

dds = DESeqDataSetFromTximport(tx_data,
                               colData = metadata,
                               design = ~ Treatment)

keep = rowSums(counts(dds)) >= 10

dds = dds[keep,]

```

#Step 3: Perform PCA analysis to ensure data integrity

```{r}

dds_norm = vst(dds)

normalized_counts = assay(dds_norm) %>%
  t()

normalized_counts_df <- as.data.frame(normalized_counts)


pca_res <- prcomp(normalized_counts_df)


pca_samples <- as.data.frame(pca_res$x)

pca_samples <- data.frame(pca_samples) %>%
  tibble::rownames_to_column("FileName") %>%
  dplyr::inner_join(metadata, by = "FileName")


plot <- ggplot(pca_samples,
  aes(
    x = PC1,
    y = PC2,
    color = Treatment
  )
) +
  geom_point(size=4) 

plot

```


# Step 4: Perform differential expression analysis

# Set the factor level so that comparison is Drug / Treatment
```{r}

dds$Treatment <- relevel(dds$Treatment, ref = "DMSO")

```


```{r}

dds_diff <- DESeq(dds)

res <- results(dds_diff)

res_df <- as.data.frame(res)

```

Step 4: Explore and export results

```{r}
summary(res)
```


```{r}

res_sig_up <- filter(res_df,
       res_df$padj < 0.05 & res_df$log2FoldChange > 2)

res_sig_down <- filter(res_df,
       res_df$padj < 0.05 & res_df$log2FoldChange < -2)

res_sig <- rbind(res_sig_up, res_sig_down)

res_sig <- rownames_to_column(res_sig, var = "Gene")


write_xlsx(res_sig, "./sig_genes.xlsx")

```


```{r}

plotMA(res)

```









